Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-26 paper.qmd
Clear the repo

```{r}
#| warning: false
#| echo: false
#| message: false
#| eval: false

@@

for (i in 1:NROW(diff_start_df)) {

@@

<!--

```{r}
#| eval: false
#| echo: false
#| warning: false

hex_full_count_df <- read_rds("data/s_curve/s_curve_hex_8.rds")

shifted_hex_coord_df <- read_rds("data/s_curve/s_curve_hex_8_shifted_hex_coord_df.rds")

bin_centroids_shift <- ggplot(data = hex_full_count_df, aes(x = c_x, y = c_y)) +
  geom_point(color = "#bdbdbd") +
  geom_point(data = shifted_hex_coord_df, aes(x = c_x, y = c_y), color = "#feb24c") +
  coord_cartesian(xlim = c(-5, 8), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

hex_grid_shift <- ggplot(data = shifted_hex_coord_df, aes(x = x, y = y)) +
  geom_polygon(fill = NA, color = "#feb24c", aes(group = polygon_id)) +
  geom_polygon(data = hex_full_count_df, aes(x = x, y = y, group = polygon_id),
               fill = NA, color = "#bdbdbd") +
  coord_cartesian(xlim = c(-5, 8), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

## Before shift
before_shift_plot <- ggplot(data = hex_full_count_df, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(aes(x = c_x, y = c_y, label = hexID), size = 2) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_equal() +
  theme_void() +
  theme(legend.position="bottom", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

## After shift
after_shift_plot <- ggplot(data = shifted_hex_coord_df, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(aes(x = c_x, y = c_y, label = hexID), size = 2) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_equal() +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| eval: false
#| echo: false
#| fig-cap: "(a) Visualization of bin centroids and (b) hexagonal grid, showcasing the configuration before (colored in gray) and after (colored in light orange) shifting the starting point by an identical amount applied in both x and y directions, with a shift value of $0.537285$."
#| label: fig-scurveshifthexgridsexp
##| out-width: 100%
#| fig-pos: H

bin_centroids_shift + hex_grid_shift +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

```{r}
#| eval: false
#| echo: false
#| fig-cap: "Hexbin plots shows the distribution of points before and after shifting the starting point of the hexagonal grid generated for UMAP applied to the training S-curve dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after shifting (identical shift applied in both x and y directions, with a shift value of $0.537285$)."
#| label: fig-scurveshifthexgrids
##| out-width: 100%
#| fig-pos: H

before_shift_plot + after_shift_plot +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

```{r}
#| eval: false
#| warning: false
#| echo: false

#MSE_df <- dplyr::bind_rows(MSE_df_1, MSE_df_2, MSE_df_3, MSE_df_4, MSE_df_5)

#MSE_df$method <- factor(MSE_df$method, levels = c("tSNE", "UMAP", "PHATE", "TriMAP", "PaCMAP"))

MSE_df <- read_rds("data/s_curve/s_curve_umap_shift_summary.rds") 
MSE_df$method <- factor(MSE_df$method, levels = c("UMAP"))

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = MSE_df |> dplyr::filter(total_mse == min(total_mse)) |> dplyr::pull(shift), linetype="dashed",
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("shift value")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = MSE_df |> dplyr::filter(total_mse == min(total_mse)) |> dplyr::pull(shift), linetype="dashed",
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("shift value")

```{r}
#| eval: false
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different shift values. In this illustration, an identical shift is applied in both the x and y directions. The Mean Squared Error (MSE) plot exhibits a global minimum at $0$, indicating that no additional shifting is necessary for the current hexagonal grid to yield a well-fitted model."
#| label: fig-diagnosticpltScurvehexbins
##| out-width: 100%
#| fig-pos: H

aic_plot + mse_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

phate_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tibble::as_tibble(PHATE_data_gau), x = "PHATE1", y = "PHATE2", hex_ratio = 50/sqrt(3)))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1",
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(PHATE_data_gau))

## Decide by looking at MSE plot
num_bins_x_phate_gau <- 60
num_bins_y_phate_gau <- 1927
hex_size_phate_gau <- 0.01
### non-empty 187

hb_obj_phate_gau <- hex_binning(data = phate_gau_scaled, x = "PHATE1", 
                               y = "PHATE2", num_bins_x = num_bins_x_phate_gau, 
                               num_bins_y = num_bins_y_phate_gau, x_start = NA, y_start = NA, 
                               buffer_x = NA, buffer_y = NA, hex_size = hex_size_phate_gau, col_start = "PHATE")

## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_phate_gau$centroids))

## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_phate_gau$hex_poly))

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_phate_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) +
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red", size = 0.1) +
#   coord_fixed()

phate_data_with_hb_id_gau <- as.data.frame(do.call(cbind, hb_obj_phate_gau$data_hb_id))

model_object_phate_gau <- fit_highd_model( training_data = training_data_gau, 
                                          nldr_df_with_id = phate_gau_scaled, 
                                          x = "PHATE1", y = "PHATE2", 
                                          num_bins_x = num_bins_x_phate_gau, 
                                          num_bins_y = num_bins_y_phate_gau, 
                                          x_start = NA, y_start = NA, 
                                          buffer_x = NA, buffer_y = NA, 
                                          hex_size = hex_size_phate_gau,
                                          is_rm_lwd_hex = FALSE, 
                                          benchmark_to_rm_lwd_hex = NA, 
                                          col_start_2d = "PHATE", 
                                          col_start_highd = "x")

df_bin_centroids_phate_gau <- model_object_phate_gau$df_bin_centroids
df_bin_phate_gau <- model_object_phate_gau$df_bin

@@

df_all_phate_gau <- dplyr::bind_cols(training_data_gau |> dplyr::select(-ID), phate_data_with_hb_id_gau)

>>>>

@@

for (i in 1:23) { #NROW(diff_start_df)

@@

```{r}
#| echo: false
#| warning: false

## Initial x and y start
initial_start_plot <- ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = df_bin_centroids, aes(x = c_x, y = c_y, label = hexID), size = 2) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_equal() +
  theme_void() +
  theme(legend.position="bottom", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

## x and y start selected by MSE

hb_obj_s_curve_n <- hex_binning(data = umap_s_curve_scaled, x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x_s_curve, 
                      num_bins_y = num_bins_y_s_curve, x_start = -0.0652, y_start = -0.195, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_s_curve, col_start = "UMAP")

## Data set with all possible centroids in the hexagonal grid
all_centroids_df_n <- as.data.frame(do.call(cbind, hb_obj_s_curve_n$centroids))

## Generate all coordinates of hexagons
hex_grid_n <- as.data.frame(do.call(cbind, hb_obj_s_curve_n$hex_poly))

## To obtain the standardise counts within hexbins
counts_df_n <- as.data.frame(do.call(cbind, hb_obj_s_curve_n$std_cts))
df_bin_centroids_n <- extract_hexbin_centroids(centroids_df = all_centroids_df_n, 
                                             counts_df = counts_df_n)

hex_grid_with_counts_n <- dplyr::left_join(hex_grid_n, counts_df_n, by = c("hex_poly_id" = "hb_id"))

new_start_plot <- ggplot(data = hex_grid_with_counts_n, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data= df_bin_centroids_n, aes(x = c_x, y = c_y, label = hexID), size = 2) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_equal() +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

hex_grid_ch_plot <- ggplot(data = hex_grid_with_counts_n, aes(x = x, y = y)) +
  geom_polygon(fill = NA, color = "#feb24c", aes(group = hex_poly_id)) +
  geom_polygon(data = hex_grid_with_counts, aes(x = x, y = y, group = hex_poly_id),
               fill = NA, color = "#bdbdbd") +
  #coord_cartesian(xlim = c(-5, 8), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

bin_centroids_ch_plot <- ggplot(data = all_centroids_df, aes(x = c_x, y = c_y)) +
  geom_point(color = "#bdbdbd") +
  geom_point(data = all_centroids_df_n, aes(x = c_x, y = c_y), color = "#feb24c") +
  #coord_cartesian(xlim = c(-5, 8), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3) 
```

```{r}
#| echo: false
#| fig-cap: "(a) Visualization of bin centroids and (b) hexagonal grid, showcasing the configuration before (colored in gray) and after (colored in light orange) shifting the starting point by an identical amount applied in both x and y directions, with a shift value of $0.537285$."
#| label: fig-scurveshifthexgridsexp
##| out-width: 100%
#| fig-pos: H

bin_centroids_ch_plot + hex_grid_ch_plot +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

```{r}
#| echo: false
#| fig-cap: "Hexbin plots shows the distribution of points before and after shifting the starting point of the hexagonal grid generated for UMAP applied to the training S-curve dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after shifting (identical shift applied in both x and y directions, with a shift value of $0.537285$)."
#| label: fig-scurveshifthexgrids
##| out-width: 100%
#| fig-pos: H

initial_start_plot + new_start_plot +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

# phate_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tibble::as_tibble(PHATE_data_gau), x = "PHATE1", y = "PHATE2", hex_ratio = 50/sqrt(3)))) |>
#   dplyr::rename(c("PHATE1" = "scaled_PHATE1",
#                   "PHATE2" = "scaled_PHATE2")) |>
#   dplyr::mutate(ID = 1:NROW(PHATE_data_gau))
# 
# ## Decide by looking at MSE plot
# num_bins_x_phate_gau <- 60
# num_bins_y_phate_gau <- 1927
# hex_size_phate_gau <- 0.01
# ### non-empty 187
# 
# hb_obj_phate_gau <- hex_binning(data = phate_gau_scaled, x = "PHATE1", 
#                                y = "PHATE2", num_bins_x = num_bins_x_phate_gau, 
#                                num_bins_y = num_bins_y_phate_gau, x_start = NA, y_start = NA, 
#                                buffer_x = NA, buffer_y = NA, hex_size = hex_size_phate_gau, col_start = "PHATE")
# 
# ## Data set with all possible centroids in the hexagonal grid
# all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_phate_gau$centroids))
# 
# ## Generate all coordinates of hexagons
# hex_grid <- as.data.frame(do.call(cbind, hb_obj_phate_gau$hex_poly))
# 
# ## To obtain the standardise counts within hexbins
# counts_df <- as.data.frame(do.call(cbind, hb_obj_phate_gau$std_cts))
# df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
#                                              counts_df = counts_df)
# # ggplot(data = hex_grid, aes(x = x, y = y)) +
# #   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
# #   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red", size = 0.1) +
# #   coord_fixed()
# 
# hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))
# 
# # ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
# #   geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
# #   geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
# #   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
# #   coord_fixed()
# 
# phate_data_with_hb_id_gau <- as.data.frame(do.call(cbind, hb_obj_phate_gau$data_hb_id))
# 
# model_object_phate_gau <- fit_highd_model( training_data = training_data_gau, 
#                                           nldr_df_with_id = phate_gau_scaled, 
#                                           x = "PHATE1", y = "PHATE2", 
#                                           num_bins_x = num_bins_x_phate_gau, 
#                                           num_bins_y = num_bins_y_phate_gau, 
#                                           x_start = NA, y_start = NA, 
#                                           buffer_x = NA, buffer_y = NA, 
#                                           hex_size = hex_size_phate_gau,
#                                           is_rm_lwd_hex = FALSE, 
#                                           benchmark_to_rm_lwd_hex = NA, 
#                                           col_start_2d = "PHATE", 
#                                           col_start_highd = "x")
# 
# df_bin_centroids_phate_gau <- model_object_phate_gau$df_bin_centroids
# df_bin_phate_gau <- model_object_phate_gau$df_bin

df_bin_centroids_phate_gau <- read_rds("data/five_gau_clusters/df_bin_centroids_phate_gau.rds")
df_bin_phate_gau <- read_rds("data/five_gau_clusters/df_bin_phate_gau.rds")

@@

# df_all_phate_gau <- dplyr::bind_cols(training_data_gau |> dplyr::select(-ID), phate_data_with_hb_id_gau)
